---
title: "Fields: dot screens, stipple and engraving"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fields: dot screens, stipple and engraving}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", dev = "ragg_png", dpi = 300, fig.retina = 1, fig.width = 89 / 25.4, fig.height = 75 / 25.4, out.width = "55%", fig.align = "center",
  fig.process = if (requireNamespace("magick", quietly = TRUE)) function(f) { magick::image_write(magick::image_quantize(magick::image_read(f), 48, dither = FALSE), f); f })   # halftone PNGs compress poorly; quantise
library(ggplot2); library(gghalftone)
vol <- data.frame(expand.grid(x = seq_len(ncol(volcano)), y = seq_len(nrow(volcano))), z = as.vector(t(volcano)))
```

`geom_halftone()` takes a gridded field (`x`, `y` and a value) and screens it directly. The examples use the Maunga Whau elevation grid from `datasets::volcano`. The area of each dot is the tone at that point. Colour scales apply to the field, and the dots inherit the colour.

## Continuous tone

```{r field}
ggplot(vol, aes(x, y, z = z)) +
  geom_halftone(aes(colour = z), angle = 45, grid = "square", gamma = 0.6) +
  with_halo(geom_contour(colour = "black", linewidth = 0.25, bins = 8)) +
  coord_equal(expand = FALSE) + labs(x = NULL, y = NULL, colour = "Elevation (m)") +
  theme_halftone(axes = "box") + theme(axis.text = element_blank(), axis.ticks = element_blank())
```

The 45 degree square lattice is the conventional screen for maps and photographs. `gamma` below 1 lifts the mid-tones.

## Engraving

A line screen draws strips along the lattice rows. Strip width follows tone. Set the halo under the contours to 0.15 mm, because a 0.09 mm halo is invisible against hatching. Where tone falls below the printable minimum, the strips break into dashes instead of getting thinner.

```{r engraving}
ggplot(vol, aes(x, y, z = z)) +
  geom_halftone(shape = "line", colour = "black", angle = 30, gamma = 1.4) +
  with_halo(geom_contour(colour = "black", linewidth = 0.2, bins = 8), width = 0.15) +
  coord_equal(expand = FALSE) + labs(x = NULL, y = NULL) +
  theme_halftone(axes = "box") + theme(axis.text = element_blank(), axis.ticks = element_blank())
```

## Illuminated contours

`with_relief()` implements illuminated contours (Tanaka 1950). Each contour segment is lit or shaded according to the direction of its slope relative to a light source, by default from the upper left. Lit segments are drawn in paper colour and shaded segments in ink. Both widen where the slope faces the light directly. Over an engraving, the paper segments cut through the hatching.

The wrapper infers which side of each contour is uphill from the neighbouring contours. `light` sets the light azimuth in degrees.

```{r relief}
ggplot(vol, aes(x, y, z = z)) +
  geom_halftone(shape = "line", colour = "black", angle = 30, gamma = 1.4) +
  with_relief(geom_contour(bins = 10), light = 315) +
  coord_equal(expand = FALSE) + labs(x = NULL, y = NULL) +
  theme_halftone(axes = "box") + theme(axis.text = element_blank(), axis.ticks = element_blank())
```

Use `with_halo()` for a line over a confidence band. Its halo is symmetric, because an interval has no lit side. Use `with_relief()` for contours over a surface.

## Stipple

`levels = 1` with blue-noise dithering (Ulichney 1993) gives a binary stipple. The other algorithms are ordered dithering (Bayer 1973) and error diffusion (Floyd and Steinberg 1976). Every dot has the same size, and dot density carries the value. Tone is capped at 55% by default, so the densest region remains a stipple rather than a filled lattice.

```{r stipple, fig.height = 62 / 25.4}
kd <- MASS::kde2d(faithful$eruptions, faithful$waiting, n = 120, lims = c(1.3, 5.6, 40, 100))
dens <- data.frame(expand.grid(x = kd$x, y = kd$y), z = as.vector(kd$z))
ggplot(dens, aes(x, y, z = z)) +
  geom_halftone(levels = 1, algorithm = "blue_noise", colour = "black") +
  with_halo(geom_contour(colour = "black", linewidth = 0.35, bins = 3)) +
  labs(x = "Eruption (min)", y = "Waiting (min)") + theme_halftone(axes = "box")
```

## Maps

`with_halftone()` on a polygon layer prints each region at the tone of its fill colour. The default continuous ramp runs from paper through ochre and red to near-black.

```{r map, fig.height = 58 / 25.4, eval = requireNamespace("maps", quietly = TRUE) && requireNamespace("mapproj", quietly = TRUE)}
val <- setNames(USArrests$Murder, tolower(rownames(USArrests)))
st <- ggplot2::map_data("state"); st$murder <- val[st$region]
ggplot(st, aes(long, lat, group = group, fill = murder)) +
  with_halftone(geom_polygon(colour = "black", linewidth = 0.2), angle = 45, grid = "square") +
  coord_map("albers", lat0 = 30, lat1 = 45) + labs(x = NULL, y = NULL, fill = "Murders\nper 100k") +
  theme_halftone(axes = "none") + theme(axis.text = element_blank())
```

## Pitch

The default pitch is 0.35 mm, or 73 lines per inch. The comparison below shows why. At 0.6 mm the screen reads as dots. At 0.25 mm it reads as a flat tint and takes five times longer to draw.

```{r ladder, fig.width = 183 / 25.4, fig.height = 45 / 25.4, out.width = "100%"}
d <- data.frame(x = seq(0, 10, length.out = 60)); d$y <- sin(d$x); d$lo <- d$y - 0.5 - 0.02 * d$x; d$hi <- d$y + 0.5 + 0.02 * d$x
one <- function(p) ggplot(d, aes(x)) + with_halftone(geom_ribbon(aes(ymin = lo, ymax = hi), fill = "black"), pitch = p) +
  with_halo(geom_line(aes(y = y), linewidth = 0.35)) + labs(x = NULL, y = NULL, title = sprintf("%.2f mm", p)) + theme_halftone() + theme(plot.title = element_text(size = 7))
patchwork::wrap_plots(lapply(c(0.6, 0.45, 0.35, 0.25), one), nrow = 1)
```

## References

- Bayer, B. E. (1973). An optimum method for two-level rendition of continuous-tone pictures. *IEEE International Conference on Communications*, 26, 11-15.
- Floyd, R. W., and Steinberg, L. (1976). An adaptive algorithm for spatial greyscale. *Proceedings of the Society for Information Display*, 17(2), 75-77.
- Tanaka, K. (1950). The relief contour method of representing topography on maps. *Geographical Review*, 40(3), 444-456. <https://doi.org/10.2307/211219>
- Ulichney, R. (1993). The void-and-cluster method for dither array generation. *Proceedings of SPIE*, 1913, 332-343. <https://doi.org/10.1117/12.152707>
- Maunga Whau elevation data: R `datasets::volcano`, digitised by Ross Ihaka from a topographic map of Auckland.
