---
title: "Halftone fills for journal figures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Halftone fills for journal figures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", dev = "ragg_png", dpi = 300, fig.retina = 1, fig.width = 89 / 25.4, fig.height = 62 / 25.4, out.width = "60%", fig.align = "center",
  fig.process = if (requireNamespace("magick", quietly = TRUE)) function(f) { magick::image_write(magick::image_quantize(magick::image_read(f), 48, dither = FALSE), f); f })   # halftone PNGs compress poorly; quantise
library(ggplot2); library(gghalftone)
```

A halftone prints tone as dots of varying size on a fixed lattice. gghalftone does this for ggplot2 fills at draw time, on a lattice with a physical pitch in millimetres, so a figure saved at 89 mm and at 183 mm carries the same screen rather than a scaled one. The intended use is the journal figure: a confidence band that stays legible in greyscale, three overlapping intervals that do not hide one another, a bar chart that a colour-blind reader can decode from the hatching alone.

## A confidence band

`with_halftone()` wraps any filled layer. The wrapped geom draws as usual; its fill is replaced by a screen, clipped to the shape, in the fill's own colour.

```{r smooth}
m <- loess(dist ~ speed, cars, span = 0.9)
nd <- data.frame(speed = seq(4, 25, length.out = 80))
pr <- predict(m, nd, se = TRUE)
nd$fit <- pr$fit; nd$lo <- pr$fit - 1.96 * pr$se.fit; nd$hi <- pr$fit + 1.96 * pr$se.fit

ggplot(nd, aes(speed)) +
  with_halftone(geom_ribbon(aes(ymin = lo, ymax = hi), fill = halftone_inks[["blue"]])) +
  with_halo(geom_line(aes(y = fit), colour = halftone_inks[["blue"]], linewidth = 0.35)) +
  geom_point(data = cars, aes(speed, dist), shape = 21, fill = "white", size = 0.8, stroke = 0.3) +
  labs(x = "Speed (mph)", y = "Stopping distance (ft)") +
  theme_halftone()
```

Three things happened without being asked for. The band's tone follows the *likelihood* of the estimate across the interval: full on the fitted line, fading to the normal density at a 95 % limit, so the fade carries the evidence rather than a look. The line over the band got a hairline of paper underneath it from `with_halo()`, so it stays readable where it crosses dots. And `theme_halftone()` set the journal register: 7 pt ticks, 8 pt titles, no gridlines, a font that is installed.

## Kaplan–Meier with three strata

Overlapping groups are the case a flat fill handles worst and a halftone handles best. Groups that share the panel are *overprinted*: their dots are woven on one lattice, so where two bands overlap both inks stay visible.

```{r km, fig.width = 89 / 25.4, fig.height = 70 / 25.4}
library(survival)
fit <- survfit(Surv(time, status) ~ ph.ecog, data = subset(lung, ph.ecog < 3))
s <- km_steps(fit); levels(s$strata) <- paste("ECOG", levels(s$strata))
cens <- km_censor(fit); levels(cens$strata) <- levels(s$strata)

ggplot(s, aes(time, group = strata)) +
  with_halftone(geom_ribbon(aes(ymin = lo, ymax = hi, fill = strata))) +
  with_halo(geom_step(aes(y = surv, colour = strata), linewidth = 0.35)) +
  geom_point(data = cens, aes(time, surv, colour = strata), shape = "|", size = 1.6) +
  scale_y_continuous(labels = scales::percent) +
  guides(colour = "none") +
  labs(x = "Days since diagnosis", y = "Overall survival", fill = NULL) +
  theme_halftone() + theme(legend.position = "inside", legend.position.inside = c(0.82, 0.86))
```

`km_steps()` turns a `survfit` into the step-shaped frame the ribbon needs; `km_censor()` and `km_risk()` give the censor marks and the number at risk. The three inks are the theme's default palette on ggplot2 4.0.

## Bars, and why colour is never the only encoding

For geoms whose groups tile the plane (bars, areas, polygons, tiles) each fill also gets its own screen angle and shape. The figure survives greyscale and the legend keys show both.

```{r bars, fig.height = 55 / 25.4}
d <- data.frame(g = factor(c("BOS", "RAS", "Mixed", "Undef."), c("BOS", "RAS", "Mixed", "Undef.")), n = c(52, 21, 14, 13))
ggplot(d, aes(g, n, fill = g)) +
  with_halftone(geom_col(width = 0.7)) +
  scale_y_continuous(expand = expansion(c(0, 0.08))) +
  labs(x = NULL, y = "Patients (%)", fill = NULL) +
  theme_halftone()
```

For a one-ink figure, hatch instead: `shape = "line"` with the `screen` aesthetic.

```{r bars_bw, fig.height = 55 / 25.4}
ggplot(d, aes(g, n, screen = g)) +
  with_halftone(geom_col(width = 0.7, fill = "black", colour = "black", linewidth = 0.25), shape = "line") +
  scale_screen_discrete(guide = "none") +
  scale_y_continuous(expand = expansion(c(0, 0.08))) +
  labs(x = NULL, y = "Patients (%)") +
  theme_halftone()
```

## Saving

`ggsave_journal()` saves at a journal column width. The extension chooses the format: PNG for proofs, TIFF for submission, PDF for a vector file in which every dot is a geometric object.

```{r save, eval = FALSE}
ggsave_journal("fig1.pdf", p, "single", height = 62)
ggsave_journal("fig1.tiff", p, "double", height = 90)
halftone_proof(p)   # true-size render plus a 4x crop of the panel centre
```

## The rules the defaults encode

- Pitch 0.35 mm (73 lines per inch). Coarser reads as a dot pattern; finer collapses into a tint.
- Nothing smaller than 0.09 mm (0.25 pt) is drawn. Light tone becomes sparse minimum-size dots, not grey pixels.
- Tone profile follows the geometry: flat where the interior is the value (bars, areas, maps), likelihood for intervals, a soft vignette for densities.
- One ink register across figure types, so a set of figures reads as a set.
- Overlaps weave; nothing translucent reaches the page. A fill's alpha becomes screen coverage.
